Abstractions for Type-Level Programming

### 2022-05-15 symposium.tex
Incorporate Dragana's feedback

The |apply| method is the entry point of our library.
It takes a regular expression and returns an instance of |Regex|.
Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's string\footnote{
\begin{diff}
The \texttt{Singleton} bound is a marker that instructs type inference to preserve unions and literal singleton types.
By default, the compiler widens those types, notably in method applications.
For instance, given \texttt{1} \texttt{::} \texttt{Nil}, Scala's default type inference yield \texttt{List[Int]} instead of the more precise \texttt{List[1]}.
To alter this behavior, one could mark \texttt{::}'s type argument with the \texttt{Singleton} bound.
\end{diff}
}.
We use that type-level string to instantiate the |Compile| type, which is our type-level analysis component.

@@

Rest assured, this is not an oversight!
Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error", on the JVM).
\begin{diff}
Even though systematic detecting of invalid regular expressions is out of the scope of our library, doing so when possible is still desirable.
\end{diff}
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.
\begin{diff}
At the time of writing, the Scala~3 compiler does not support the customization of match type errors.
\end{diff}

@@

The Haskell ecosystem contains several packages for regular expressions, with various levels of static safety.
The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
The main drawback of this solution lies in the library's learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.
\begin{diff}
Tyre~\citep{radanne2020tyre} is an OCaml library that also uses combinators to define type-safe regular expressions.
\end{diff}

>>>>

The |apply| method is the entry point of our library.
It takes a regular expression and returns an instance of |Regex|.
Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's |String|\footnote{
\begin{diff}
The \texttt{Singleton} bound is a marker that instructs type inference to preserve unions and literal singleton types.
By default, the compiler widens those types, notably in method applications.
For instance, given \texttt{1} \texttt{::} \texttt{Nil}, Scala's default type inference yields \texttt{List[Int]} instead of the more precise \texttt{List[1]}.
To alter this behavior, one could mark \texttt{::}'s type argument with the \texttt{Singleton} bound.
\end{diff}
}.
We use that type-level string to instantiate the |Compile| type, which is our type-level analysis component.

@@

Rest assured, this is not an oversight!
Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error", on the JVM).
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.
\begin{diff}
At the time of writing, the Scala~3 compiler does not support the customization of match type errors.
The systematic detection of invalid regular expressions is out of the scope of this paper.
\end{diff}

@@

The Haskell ecosystem contains several packages for regular expressions, with various levels of static safety.
The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
\begin{diff}
Tyre~\citep{radanne2020tyre} is an OCaml library that also uses combinators to define type-safe regular expressions.
The main drawback of these solutions is the learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.
\end{diff}
